---
title: "The Music of the Solar System"
subtitle: "An Interactive Journey Through Planetary Data"
author: "Data Science Presentation"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: false
    css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(plotly)
library(dplyr)
```

## Welcome to the Solar System {.slide-title}

> *"The cosmos is within us. We are made of star-stuff."*  
> — Carl Sagan

This presentation explores **planetary data** through interactive Plotly visualizations.  
Navigate the slides to discover patterns in our solar neighborhood.

**Created:** `r format(Sys.Date(), '%B %d, %Y')`

---

## The Data {.data-slide}

We use NASA's planetary constants to explore relationships between:

- 🪐 **Orbital Period** — how long each planet takes to orbit the Sun
- 📏 **Distance from Sun** — semi-major axis in Astronomical Units (AU)  
- 🌍 **Planet Mass** — relative to Earth
- 🌡️ **Mean Temperature** — surface temperature in Kelvin
- 🔵 **Planet Size** — radius relative to Earth

```{r planet-data}
planets <- data.frame(
  name        = c("Mercury","Venus","Earth","Mars",
                  "Jupiter","Saturn","Uranus","Neptune"),
  distance_au = c(0.387, 0.723, 1.000, 1.524,
                  5.203, 9.537, 19.19, 30.07),
  period_days = c(87.97, 224.7, 365.25, 686.97,
                  4332.6, 10759, 30687, 60190),
  mass_earth  = c(0.055, 0.815, 1.000, 0.107,
                  317.8, 95.16, 14.54, 17.15),
  radius_km   = c(2439, 6052, 6371, 3390,
                  69911, 58232, 25362, 24622),
  temp_k      = c(440, 737, 288, 210,
                  165, 134, 76, 72),
  color       = c("#b5b5b5","#e8cda0","#4a9af5","#c1440e",
                  "#c9a96e","#e8d5a3","#72d7e8","#4169e1"),
  type        = c("Rocky","Rocky","Rocky","Rocky",
                  "Gas Giant","Gas Giant","Ice Giant","Ice Giant")
)
```

---

## Interactive Planetary Explorer {.plotly-slide}

```{r main-plot, fig.width=10, fig.height=5.5}
fig <- plot_ly(
  planets,
  x = ~distance_au,
  y = ~period_days,
  size = ~radius_km,
  color = ~type,
  colors = c("Rocky" = "#e8634a", "Gas Giant" = "#f0b429",
             "Ice Giant" = "#38bdf8"),
  text = ~paste0(
    "<b>", name, "</b><br>",
    "Distance: ", distance_au, " AU<br>",
    "Orbital Period: ", formatC(period_days, format="f", digits=0), " days<br>",
    "Mass: ", mass_earth, "× Earth<br>",
    "Radius: ", formatC(radius_km, big.mark=",", format="d"), " km<br>",
    "Avg Temp: ", temp_k, " K"
  ),
  hoverinfo = "text",
  type = "scatter",
  mode = "markers",
  marker = list(
    sizemode = "diameter",
    sizeref = 2,
    opacity = 0.85,
    line = list(color = "white", width = 1.5)
  )
) %>%
  layout(
    title = list(
      text = "Planetary Orbital Period vs Distance from the Sun",
      font = list(size = 18, color = "#1e293b", family = "Georgia")
    ),
    xaxis = list(
      title = "Distance from Sun (Astronomical Units)",
      type  = "log",
      gridcolor = "#e2e8f0",
      tickfont  = list(color = "#64748b"),
      titlefont = list(color = "#334155")
    ),
    yaxis = list(
      title = "Orbital Period (Earth Days)",
      type  = "log",
      gridcolor = "#e2e8f0",
      tickfont  = list(color = "#64748b"),
      titlefont = list(color = "#334155")
    ),
    legend = list(
      title = list(text = "<b>Planet Type</b>"),
      font  = list(color = "#334155")
    ),
    plot_bgcolor  = "#f8fafc",
    paper_bgcolor = "#f8fafc",
    hoverlabel = list(
      bgcolor   = "#1e293b",
      font      = list(color = "white", size = 13)
    ),
    annotations = list(
      list(
        x = log10(1), y = log10(365),
        text = "🌍 Earth", showarrow = FALSE,
        font = list(size = 12, color = "#334155"),
        xshift = 20, yshift = 10
      )
    )
  )

fig
```

*Hover over each planet · Bubble size = planet radius · Axes are log-scaled*

---

## Kepler's Third Law — Visualised {.kepler-slide}

The log-log plot reveals **Kepler's Third Law**: $T^2 \propto a^3$

```{r kepler, fig.width=9, fig.height=4.5}
kepler <- planets %>%
  mutate(
    log_dist   = log10(distance_au),
    log_period = log10(period_days / 365.25),  # convert to years
    label      = name
  )

fit <- lm(log_period ~ log_dist, data = kepler)
x_seq <- seq(-0.5, 1.55, length.out = 100)
y_fit <- predict(fit, newdata = data.frame(log_dist = x_seq))

fig2 <- plot_ly() %>%
  add_trace(
    data = kepler,
    x = ~log_dist, y = ~log_period,
    type = "scatter", mode = "markers+text",
    text = ~label, textposition = "top center",
    marker = list(
      color = planets$color, size = 14,
      line = list(color = "white", width = 1.5)
    ),
    hovertext = ~paste0(name, "<br>log(a) = ", round(log_dist,3),
                        "<br>log(T) = ", round(log_period,3)),
    hoverinfo = "text",
    name = "Planets"
  ) %>%
  add_trace(
    x = x_seq, y = y_fit,
    type = "scatter", mode = "lines",
    line = list(color = "#6366f1", width = 2, dash = "dash"),
    name = paste0("Fit (slope = ", round(coef(fit)[2],2), ")")
  ) %>%
  layout(
    title = list(
      text = "Kepler's Third Law: log(Period) vs log(Distance)",
      font = list(size = 16, color = "#1e293b", family = "Georgia")
    ),
    xaxis = list(title = "log₁₀(Distance in AU)", gridcolor = "#e2e8f0",
                 titlefont = list(color="#334155"), tickfont=list(color="#64748b")),
    yaxis = list(title = "log₁₀(Period in Years)", gridcolor = "#e2e8f0",
                 titlefont = list(color="#334155"), tickfont=list(color="#64748b")),
    plot_bgcolor  = "#f8fafc",
    paper_bgcolor = "#f8fafc",
    legend = list(font = list(color = "#334155")),
    hoverlabel = list(bgcolor="#1e293b", font=list(color="white", size=13))
  )

fig2
```

The best-fit slope is **≈ 1.5**, confirming $T \propto a^{3/2}$ — a perfect power law hiding in our solar system.

---

## Temperature vs Distance {.temp-slide}

```{r temp-plot, fig.width=9, fig.height=4.5}
fig3 <- plot_ly(
  planets,
  x = ~distance_au,
  y = ~temp_k,
  type = "scatter",
  mode = "markers+lines",
  marker = list(
    color = ~color, size = ~log(radius_km)*2.5,
    line = list(color="white", width=1.5),
    opacity = 0.9
  ),
  line = list(color = "#94a3b8", width = 1, dash = "dot"),
  text = ~paste0("<b>", name, "</b><br>",
                 round(distance_au, 2), " AU · ",
                 temp_k, " K (", round(temp_k - 273, 0), "°C)"),
  hoverinfo = "text"
) %>%
  add_annotations(
    x = planets$distance_au,
    y = planets$temp_k,
    text = planets$name,
    showarrow = FALSE,
    yshift = 14,
    font = list(size = 11, color = "#334155")
  ) %>%
  layout(
    title = list(
      text  = "Surface Temperature Plummets with Distance",
      font  = list(size = 16, color = "#1e293b", family = "Georgia")
    ),
    xaxis = list(title = "Distance from Sun (AU)", type = "log",
                 gridcolor = "#e2e8f0",
                 titlefont = list(color="#334155"), tickfont=list(color="#64748b")),
    yaxis = list(title = "Mean Temperature (Kelvin)", gridcolor = "#e2e8f0",
                 titlefont = list(color="#334155"), tickfont=list(color="#64748b")),
    shapes = list(
      list(type="line", x0=0.3, x1=35, y0=273, y1=273,
           line=list(color="#22c55e", width=1.5, dash="dash"))
    ),
    annotations = list(
      list(x=1.2, y=273, text="0°C (Water freezes)", showarrow=FALSE,
           font=list(color="#22c55e", size=11), xanchor="left")
    ),
    plot_bgcolor  = "#f8fafc",
    paper_bgcolor = "#f8fafc",
    showlegend    = FALSE,
    hoverlabel    = list(bgcolor="#1e293b", font=list(color="white", size=13))
  )

fig3
```

*Note: Venus is hotter than Mercury due to its extreme greenhouse effect.*

---

## Key Takeaways

### What the data shows us:

1. **Kepler's Third Law holds perfectly** — $T^2 = a^3$ gives slope 1.5 on log-log axes, a relationship discovered in 1619.

2. **Temperature follows an inverse power law** with distance — the outer solar system is brutally cold (Neptune: −201°C).

3. **Size doesn't follow a simple gradient** — the gas giants dwarf the inner rocky planets by orders of magnitude.

4. **Plotly makes exploration effortless** — zoom, hover, and filter to find patterns yourself.

---

**Data sources:** NASA Planetary Fact Sheets  
**Tools:** R · Plotly · R Markdown · ioslides  
**Created:** `r format(Sys.Date(), '%B %d, %Y')`

*Upload this HTML to GitHub Pages, RPubs, or Neocities to share publicly.*
